---
title: "Exercise 1"
author: "Roi Aharon, Hila Livnat, Daniel Erez"
date: "27 March 2025"
output:
html_document:
code_folding: hide
editor_options: 
markdown: 
wrap: sentence
---

```{r}
install.packages("ggplot2")
install.packages("GGally")
library(ggplot2)
library(GGally)
```

**Section A – File Reading:**

```{r}
#Qa
dir_name <- "/Users/danielerez/data"
file_C <- sprintf("%s/%s", "data", "chr1_str_50M_70M.rda")
data_C = load(file_C)
region_line_C <- strsplit(chr1_str_50M_70M, split = "")[[1]]
bin_size <- 2000
regionC_factored = factor(region_line_C, c("A","T","C","G"))
counts_C <- cell_counts(regionC_factored, cell_size = bin_size)


```

```{r}
# table(region_line_C)
```

```{r}
# regionA_factored = factor(region_line_A, c("A","T","C","G"))
# regionB_factored = factor(region_line_B, c("A","T","C","G"))
# round(table(regionC_factored), digits = -4) /10^6
```

**Section B – A function that receives a genomic region and calculates the number of occurrences of each base in that region:**

A function that takes a region in the genome and calculates the number of appearances of each base (A, T, C, G) in that region.

```{r}
#Q b
# Receives region and cell size and summarizes number of bases in each cell
cell_counts = function(region, cell_size = 2000) {
  n = length(region)
  n_cells = ceiling(n / cell_size)
  
  res = data.frame(
    cell =  integer(n_cells),
    A = integer(n_cells),
    T = integer(n_cells),
    C = integer(n_cells),
    G = integer(n_cells)
  )
  
  
  # counting the number of appearances
  for (i in 1:n_cells) {
    start_idx = (i-1) * cell_size + 1
    end_idx = min(i * cell_size, n)
    cell_segment = region[start_idx:end_idx]
    base_counts = table(cell_segment)
    
    res$cell[i] = i
    res$A[i] <- ifelse("A" %in% names(base_counts), base_counts["A"], 0)
    res$T[i] <- ifelse("T" %in% names(base_counts), base_counts["T"], 0)
    res$C[i] <- ifelse("C" %in% names(base_counts), base_counts["C"], 0)
    res$G[i] <- ifelse("G" %in% names(base_counts), base_counts["G"], 0)
  }
  return (res)
} 
```

```{r}
```

**Section C – In each histogram, you can see the distribution of each of the bases in the chromosome:**

You can see that A and T are relatively close to a normal distribution, while the others look like a normal distribution with a right-hand tail.

```{r}
#Q c
# histogram
#create historgram dist for each chromozon
hist_colors <- c("red", "blue", "green", "purple")
chroms <- c("A","T", "C", "G")
for (j in seq_along(chroms)){
  N = length(regionC_factored)
  n_bins = ceiling(N/bin_size)
  C_count = numeric(N/bin_size)
  for (i in 1:n_bins){
    start_bin = 1+ (i-1)*bin_size
    end_bin = i*bin_size
    C_count[i] = sum(regionC_factored[start_bin:end_bin] ==chroms[j])
  }
  
  hist(C_count,
       main = paste("Distribution of", chroms[j], "per", bin_size, "bases"),
       xlab = paste("Count of", chroms[j]),
       ylab = "Frequency",
       col = hist_colors[j],
       border = "darkblue"
      )
  rug(C_count, color="orange")
}
```

**Section D –**

In the above plots, we see the frequency of specific bases in cells of size 2000 relative to their position on the chromosome. You can observe that the lines follow the same trend line (the dashed black line). There's a slight increase or decrease around cell 2000, but overall, the trend is consistent across all lines, and the frequency forms a sort of straight line.

```{r}
#Q d
# Set up 2x2 plotting area
# ylim_range <- c(0, 1)
par(mfrow = c(2, 2))
# 
# # Plot A
# plot(counts_C$cell, counts_C$A/bin_size, type = "l", lty = 1, lwd = 2,
#      col = hist_colors[1],
#      xlab = "Chromosome Position (Cell Index)",
#      ylab = "Frequency",
#      main = "Frequency of A",
#      ylim = ylim_range)
# 
# # Plot T
# plot(counts_C$cell, counts_C$T/bin_size, type = "l", lty = 1, lwd = 2,
#      col = hist_colors[2],
#      xlab = "Chromosome Position (Cell Index)",
#      ylab = "Frequency",
#      main = "Frequency of T", 
#      ylim = ylim_range)
# 
# # Plot C
# plot(counts_C$cell, counts_C$C/bin_size, type = "l", lty = 1, lwd = 2,
#      col = hist_colors[3],
#      xlab = "Chromosome Position (Cell Index)",
#      ylab = "Frequency",
#      main = "Frequency of C",
#      ylim = ylim_range)
# 
# # Plot G
# plot(counts_C$cell, counts_C$G/bin_size, type = "l", lty = 1, lwd = 2,
#      col = hist_colors[4],
#      xlab = "Chromosome Position (Cell Index)",
#      ylab = "Frequency",
#      main = "Frequency of G", 
#      ylim = ylim_range)
# 

# Create boxplot for base frequencies across all cells
# boxplot(counts_C[, c("A", "T", "C", "G")],
#         col = hist_colors,
#         ylab = "Frequency",
#         xlab = "Base",
#         main = "Distribution of Base Frequencies Across Cells")

plot_base_frequency <- function(base, data = counts_C, color = "blue", trend_col = "black") {
  # Check if the base column exists
  if (!(base %in% names(data))) {
    stop(paste("Base", base, "not found in data"))
  }
  
  # Extract cell positions and base frequencies
  x <- data$cell
  y <- data[[base]]
  
  # Plot raw frequency data
  plot(x, y, type = "l", lwd = 2,
       col = color,
       xlab = "Cell Number (Index)",
       ylab = paste("Frequency of", base),
       main = paste("Frequency of", base, "Across Cells"))
  
  # Add loess-smoothed trend line
  trend <- loess(y ~ x, span = 0.1)  # adjust `span` for more/less smoothing
  lines(x, predict(trend), col = trend_col, lwd = 2, lty = 2)
}

for (j in seq_along(chroms)){
  plot_base_frequency(chroms[j], color = hist_colors[j], trend_col = "black")

}

```

**Section E – In the following plots, we can see the frequency for every pair of bases:**

While the general trend shows that as the quantity of one base increases, the quantity of another decreases, there’s an interesting exception when comparing A to T and C to G. For these two pairs, as the amount of one base increases, so does the amount of the other. Cells with relatively high quantities of C and G are mostly located in the two central quarters of the chromosome, while cells with relatively high quantities of A and T are mainly found in the first three quarters and are almost entirely absent at the end of the chromosome.

Additionally, we marked the top 0.1% most extreme outlier cells in red. When examining their locations, we found that across all plots, most of these cells are concentrated between approximately cell 1000 and cell 2000, and the final cell also appears in every graph. Upon printing these specific cells, we noticed they contain many missing values (NAs), suggesting that the outlier behavior likely stems from a lack of information about the bases.



```{r}
# # Plot all bases together
# matplot(counts_C$cell, counts_C[, c("A", "T", "C", "G")], type = "l", lty = 1, lwd = 2,
#         col = hist_colors,
#         xlab = "Chromosome Position (Cell Index)",
#         ylab = "Base Frequency",
#         main = "Base Frequencies Across Chromosome")
# 
# legend("topright", legend = c("A", "T", "C", "G"),
#        col = hist_colors, lty = 1, lwd = 2)
# Set up 2x2 plotting area
# Base names and number of unique combinations
bases <- c("A", "C", "G", "T")


n_cells <- nrow(counts_C)
divied_to = 4

# Create a bin index from 1 to 10
counts_C$cell_bin <- cut(counts_C$cell, breaks = divied_to, labels = FALSE)
bin_colors <- rainbow(divied_to)

# Set up plotting area: 2 rows x 3 columns
par(mfrow = c(2, 3))

# Loop over unique base combinations
for (i in 1:3) {
  for (j in (i+1):4) {
    x <- counts_C[[bases[i]]]
    y <- counts_C[[bases[j]]]

    # Calculate Mahalanobis distance
    data_pair <- cbind(x, y)
    md <- mahalanobis(data_pair, colMeans(data_pair), cov(data_pair))

    # Define a cutoff for outliers (e.g. top 1%)
    cutoff <- quantile(md, 0.999)
    outlier_idx <- which(md > cutoff)

    # Plot
    plot(x, y,
         # col = bin_colors[counts_C$cell_bin],
         pch = 20, cex = 0.5,
         xlab = paste("Freq of", bases[i]),
         ylab = paste("Freq of", bases[j]),
         main = paste(bases[i], "vs", bases[j]))

    # Highlight outliers
    points(x[outlier_idx], y[outlier_idx], col = "red", pch = 4, cex = 0.8)

    # Extract cell numbers
    outlier_cells <- counts_C$cell[outlier_idx]
    print(paste("Outliers in", bases[i], "vs", bases[j], ":"))
    print(outlier_cells)
  }
}

# # Loop over unique base combinations
# for (i in 1:3) {
#   for (j in (i+1):4) {
#     plot(counts_C[[bases[i]]], counts_C[[bases[j]]],
#          col = bin_colors[counts_C$cell_bin],
#          pch = 20, cex = 0.5,
#          xlab = paste("Freq of", bases[i]),
#          ylab = paste("Freq of", bases[j]),
#          main = paste(bases[i], "vs", bases[j]))
#   }
# }
# 
# # Assuming you already ran this:
# threshold <- 0.3
# # Get high-GC cell rows
# high_GC <- counts_C[counts_C$G/bin_size  > threshold & counts_C$C/bin_size > threshold, ]
# 
# high_AT<- counts_C[counts_C$A/bin_size > threshold & counts_C$T/bin_size > threshold, ]
# 
# print(high_AT)
# # # Loop through each high-GC cell and print the sequence
# # for (i in seq_len(nrow(high_GC))) {
# #   cell_index <- high_GC$cell[i]  # Get cell number (index)
# #   
# #   # Calculate start and end positions in chr22_line
# #   start_pos <- (cell_index - 1) * bin_size + 1
# #   end_pos <- min(cell_index * bin_size, length(region_line_C))  # Avoid overflow
# #   
# #   # Extract the sequence
# #   seq_segment <- region_line_C[start_pos:end_pos]
# #   
# #   # Print header and sequence
# #   cat(paste0("Cell ", cell_index, " (", start_pos, "-", end_pos, "):\n"))
# #   cat(paste(seq_segment, collapse = ""), "\n\n")
# # }
# 
# # # Create a factor with range labels
# # cut_ranges <- cut(counts_C$cell, breaks = 10)
# # 
# # # Print out the color mapping
# # for (i in 1:10) {
# #   cat(paste("Color", i, "=", bin_colors[i], "→ Cell range:", levels(cut_ranges)[i]), "\n")
# # }
# 

```
```{r}
# n_bins = ceiling(N/bin_size)
# let_count = matrix(nr = n_bins,nc = 5)
# letters = c("A","C","G","T")
# for (i in 1:n_bins){
#   start_bin = 1+ (i-1)*bin_size
#   end_bin = min(N-1, i*bin_size)
#   if (i > 10000){
#     print(start_bin)
#     print(end_bin)
#     print(N)
#   }
#   for (let in 1:4) {
#     let_count[i,let] = sum(region_line_C[start_bin:end_bin] ==letters[let],na.rm=TRUE)
#     
#   }
# }
# par(mfcol = c(4,1),mar = c(1,0,0,0))
# for (j in 1:4){
#   plot((1:n_bins),
#      let_count[,j]/bin_size,
#      pch =20, 
#      cex = 0.5, 
#      xlab = "Genomic Position (bp)",
#      ylab = paste(letters[j], " Content (%)"),
#      main = paste("Percentage of ", letters[j], " Across Genome Bins"),
#      xaxt = "n")  # hide default X-axis ticks
# 
#   # Set tick step to 1 kb
#   tick_step <- 1e3  # 1000 bp
#   
#   # Add custom X-axis labels every 1 kb
#   axis(1,
#        at = seq(0, max(n_bins * bin_size), by = tick_step),
#        labels = seq(0, max(n_bins * bin_size), by = tick_step))
#   
# }

```
```{r}
par(mfcol = c(4,1),mar = c(1,0,0,0))
for (i in 1:4){
  plot((1:n_bins)*bin_size,let_count[,i]/bin_size,pch =20, cex = 0.5, xlab = "Location")
}
```


```{r}
# Set up empty plot space
plot(1:divied_to, rep(1, divied_to), type = "n", axes = FALSE, xlab = "", ylab = "",
     main = "Color Legend: Cell Number Ranges")

# Draw colored rectangles with text
for (i in 1:10) {
  rect(i - 0.4, 0.8, i + 0.4, 1.2, col = bin_colors[i], border = "black")
  text(i, 0.6, labels = paste0("(", (i-1)*1000, "-", i*1000, "]"), cex = 0.8)
}
```

**Section F – According to the boxplot above, we would divide the groups as follows: {A, T} and {C, G}:**

You can see that the frequency of the first group is much higher, but also with greater variance, while the frequency of the second group is lower, but with less variance. Therefore, we believe this is the logical grouping of bases that behave similarly.

```{r}
#Q f
# Create boxplot for base frequencies across all cells
boxplot(counts_C[, c("A", "T", "C", "G")],
        col = hist_colors,
        ylab = "Frequency",
        xlab = "Base",
        main = "Distribution of Base Frequencies Across Cells")
```
