---
title: "ex4"
author: "Roei Aharon, Hila Livnat, Daniel Erez"
output:
  html_document:
    code_folding: hide
editor_options: 
markdown: 
wrap: sentence
date: "2025-05-01"

---
```{r}
library('data.table')
library(ggplot2)
library(mgcv)
library(zoo)
library(dplyr)
source("helper_functions.R")


# Set directory
rda_dir <- "c1_bases"
rda_files <- list.files(rda_dir, pattern = "\\.rda$", full.names = TRUE)
bin_size_GC = 5000


# Function to load string from .rda and convert to character vector
extract_region_line <- function(file) {
  env <- new.env()
  load(file, envir = env)
  obj_names <- ls(env)

  # Optional: narrow down to a specific expected variable, if needed
  string_var <- env[[obj_names[1]]]  # assumes only 1 object or you want the first one

  if (!is.character(string_var) || length(string_var) != 1) {
    stop(paste("Expected a single character string in:", file))
  }

  strsplit(string_var, split = "")[[1]]
}

# Apply across all files
region_line_parts <- lapply(rda_files, extract_region_line)

# Combine all character vectors into one
region_line_C <- unlist(region_line_parts)

# Check result
# print(head(region_line_C))
# cat("Total length:", length(region_line_C), "\n")

cell_counts = function(region, cell_size = 2000) {
  n = length(region)
  n_cells = ceiling(n / cell_size)
  
  # res = data.frame(
  #   cell =  integer(n_cells),
  #   # A = integer(n_cells),
  #   # T = integer(n_cells),
  #   GC = integer(n_cells),
  #   # G = integer(n_cells)
  # )
  # 
  res =integer(n_cells)
  
  # counting the number of appearances
  for (i in 1:n_cells) {
    start_idx = (i-1) * cell_size + 1
    end_idx = min(i * cell_size, n)
    cell_segment = region[start_idx:end_idx]
    base_counts = table(cell_segment)
    
    # res$A[i] <- ifelse("A" %in% names(base_counts), base_counts["A"], 0)
    # res$T[i] <- ifelse("T" %in% names(base_counts), base_counts["T"], 0)
    res[i] <- ifelse("C" %in% names(base_counts), base_counts["C"], 0) + ifelse("G" %in% names(base_counts), base_counts["G"], 0)
    # res$G[i] <- ifelse("G" %in% names(base_counts), base_counts["G"], 0)
  }
  return (res)
}

GC <- cell_counts(region_line_C, bin_size_GC)

reads_file <- 'TCGA-13-0723-10B_lib1_all_chr1.forward'
chr1_reads = fread(reads_file) 
colnames(chr1_reads) = c("Chrom","Loc","FragLen")
Num_cells = max(chr1_reads$Loc)
covers <- getReadLineFast(chr1_reads$Loc, min(chr1_reads$Loc), Num_cells)

bin_nums <- ceiling(Num_cells/bin_size_GC)
not_filtered_bin_covers <- numeric(bin_nums -1)
for (k in 1:bin_nums-1){
  not_filtered_bin_covers[k] = sum(covers[bin_size_GC*(k-1)+1 : bin_size_GC*k])
}

# Step 1: Remove NAs
valid_indices <- which(!is.na(not_filtered_bin_covers) & seq_along(not_filtered_bin_covers) <= length(GC))
GC <- GC[valid_indices]

reads <- na.omit(not_filtered_bin_covers)
```
```{r}
model_data <- data.frame(GC = GC, reads = reads)


# Assuming you have a data.frame named model_data with columns GC and reads

ggplot(model_data, aes(x = GC, y = reads)) +
  geom_point(alpha = 0.3, size = 1, color = "steelblue") +  # Raw data points
  geom_smooth(method = "gam", formula = y ~ s(x), color = "darkred", se = TRUE, linewidth = 1.2) +
  labs(
    title = "Trend Between GC and Bin Covers (Bin Size = 5000)",
    x = "GC Content",
    y = "Bin Covers",
    caption = "The line represents a smooth GAM regression"
  ) +
  theme_minimal(base_size = 14)
```

```{r}
gam_model <- gam(reads ~ s(GC), data = model_data)
model_data <- model_data %>% mutate(fitted = predict(gam_model),
         residuals = residuals(gam_model))

ggplot(model_data, aes(x = GC, y = residuals)) +
  geom_point(alpha = 0.3, color = "darkblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residuals vs Fitted Values",
    x = "GC count",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 14)
```
The residuals vs. fitted values plot reveals a noticeable bias in the range where fitted values are approximately between 225 and 275. In this region, the residuals exhibit a distinct vertical clustering with a wider spread and a tendency to deviate from zero more consistently than in other parts of the plot. This pattern suggests that the model is systematically under- or over-predicting in this range, rather than producing residuals that are symmetrically distributed around zero as expected in a well-fitted model. The dense vertical band implies that the model struggles to capture the true relationship between GC content and read count in this segment, likely due to insufficient flexibility or unaccounted non-linearity in the data. This local misfit indicates a bias in the model’s predictions for lower GC content levels, highlighting the need for either a more flexible model or a targeted correction in this range

```{r}
# Filter the data
filtered_data <- model_data %>% filter(GC >= 1500)

# Fit the GAM model on filtered data
gam_model_filtered <- gam(reads ~ s(GC), data = filtered_data)

# Add fitted values and residuals to filtered data only
filtered_data <- filtered_data %>%
  mutate(fitted = predict(gam_model_filtered),
         residuals = residuals(gam_model_filtered))


ggplot(model_data, aes(x = GC, y = reads)) +
  geom_point(alpha = 0.3, color = "grey60") +  # All data
  geom_line(data = filtered_data, aes(x = GC, y = fitted), color = "red", size = 1.2) +
  labs(title = "Regression Fit for GC >= 1500", x = "GC count", y = "Bin Covers") +
  theme_minimal(base_size = 14)
# 
# ggplot(filtered_data, aes(x = GC, y = residuals)) +
#   geom_point(alpha = 0.4, color = "darkblue") +
#   geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
#   labs(
#     title = "Residuals vs Fitted (GC ≥ 1500)",
#     x = "Fitted Values",
#     y = "Residuals"
#   ) +
#   theme_minimal(base_size = 14)
```

```{r}
mean_residual_low  <- mean(filtered_data$residuals[filtered_data$GC <= 3000], na.rm = TRUE)
mean_residual_high <- mean(filtered_data$residuals[filtered_data$GC > 3000],  na.rm = TRUE)

ggplot(filtered_data, aes(x = GC, y = residuals)) +
  geom_point(alpha = 0.3, color = "green") +
  geom_hline(yintercept = mean_residual_low, color = "purple", linetype = "dotted", size = 1) +
  geom_hline(yintercept = mean_residual_high, color = "darkblue", linetype = "dotted", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = min(filtered_data$GC), y = mean_residual_low + 10,
           label = paste0("Mean GC ≤ 3000: ", round(mean_residual_low, 2)),
           color = "purple", hjust = 0) +
  annotate("text", x = max(filtered_data$GC), y = mean_residual_high + 10,
           label = paste0("Mean GC > 3000: ", round(mean_residual_high, 2)),
           color = "darkblue", hjust = 1) +
  labs(
    title = "Residuals vs GC >= 1500 with Mean Residuals by Segment",
    x = "GC Count",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 14)
```

As seen in the first graph, the center of the distribution occurs when GC is greater than 1500. Therefore, we focused only on that part and performed a regression on it. We then examined the residuals for this section alone and observed that up to GC = 3000, the residuals behave nicely and are concentrated around 0. However, beyond that point, the residuals are more dispersed and not well-centered around 0. So, we calculated the average residual up to 3000 and found it to be approximately 0, as expected. But above 3000, the average residual is 0.22, indicating that the residuals are not well-centered around 0 in that range and there is still some bias.




